### 2025-08-15 docs/Homeownership_BankOfMomAndDad.tex
Ready for third submission!

\newcommand{\E}{\mathbb{E}}
\newcommand{\mc}{\mathcal}

\title{{\LARGE Illiquid Homeownership and the \\ Bank of Mom and Dad}\thanks{Federal Reserve Board of Governors. E-mail: eirik.e.brandsaas@frb.gov. I have benefited from the comments of Dean Corbae, Erwan Quintin, Ananth Seshadri, Corina Mommaerts, Corina Boar, Manuel Amador, Hamish Low, Patrick Moran, Dionissi Aliprantis, David Sturrock, Mike Eriksen, Aradhya Sood, Margaret Jacobson, William Gamber, Claudia Sahm, and Joel McMurry, as well as seminar participants Oxford Sa{\"i}d, Federal Reserve Board, Cal State Fullerton, Boston Fed, ECARES/Université libre de Bruxelles, Santa Clara, University of Lugano, NYU Furman, Norges Bank, BI Norwegian Business School, UW-Madison, and conference participants at NASEMS, AREUEA National, ReCapNet, Minneapolis Fed/OIGI, UEA EU, CFPB, CEF, SAET, Oxford NuCamp Macro Workshop, YES 2020 at UPenn, and Minnesota-Wisconsin Macro/International Workshop. Code and replication instructions are available at the \href{https://github.com/eirikbrandsaas/HomeownershipBankMomDad.jl}{author's  website and Github}.\newline \textit{Disclaimer: }The views expressed in this paper are solely the responsibility of the author and should not be interpreted as reflecting the views of the Board of Governors of the Federal Reserve System or of anyone else associated with the Federal Reserve System.}}
\author{Eirik Eylands Brandsaas}
\date{{\today }}

@@

\maketitle
\begin{abstract}
{\begin{spacing}{1.2}
Housing is the largest asset in U.S. household portfolios, and first-time homebuyers increasingly rely on parental transfers. This paper quantifies the contribution of parental transfers to the homeownership rate of young households. I build and estimate a life-cycle overlapping generations model with housing, where adult children and parents interact without commitment. I find that parental transfers account for 14 percentage points (29\%) of young households' homeownership. Transfers from wealthy parents not only help households overcome borrowing constraints, but also help sustain homeownership, mitigating the drawbacks of illiquidity. Surprisingly, policies lowering entry barriers to homeownership generally increase the reliance on parental wealth, whereas increased liquidity reduces it. Finally, I show that children of wealthy parents strategically use the illiquidity of housing as a commitment device to encourage transfers, resulting in a preference for illiquidity.
\end{spacing}}
\vspace{0.5cm}
\noindent \textbf{JEL:} D14, D15, E21, G51, R21

@@

\section{Introduction}
Housing is the largest asset in U.S. household portfolios. While homeownership is seen as central to the ``American Dream'' \citep{goodman2018homeownership}, households face different barriers to ownership. Between 2009 and 2016, around 30\% of American first-time homebuyers received direct parental assistance for their down payment.\footnote{Source: Survey of Household Economic Decisionmaking (see Section \ref{sec:data}).} This assistance highlights disparities in access, as only those with sufficiently wealthy parents can expect substantial transfers.

@@

To quantify the impact of parental transfers on homeownership, I then combine two distinct models: housing models, which ignore parental transfers, and models of parental transfers, which ignore housing. I build a life-cycle overlapping generations model with altruistic parents and a rent-or-own decision. Parents and children interact without commitment through both inter-vivos transfers and end-of-life bequests. In models of altruism without housing (e.g., \citealp{Altonji1997a,Barczyk2014}), the transfer motive is to increase the child's consumption when they are borrowing constrained, thus having high marginal utility of wealth. The inclusion of a frictional housing market generates two new transfer motives. First, transfers can directly alleviate credit constraints, enabling wealthier parents to help their children meet down-payment requirements. Second, future transfers provide partial insurance against future income shocks, reducing risks associated with the large illiquid investment of homeownership. Moreover, the illiquidity of housing serves as a commitment device enabling children to secure future transfers. These mechanisms differ from those examined by \citet{Barczyk2020a}, who study how elderly parents’ homeownership decisions shape nursing‑home entry, care arrangements, and intergenerational transfers. As in all transfer models, the expectation of future transfers reduces children's saving motives, potentially lowering the homeownership rate.

@@

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Study of Income Dynamics (PSID). To quantify the importance of transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 13 percentage points (27\%) of the homeownership rate among young households. In one adaption of my model, I quantify the contribution of parental transfers to the Black-White homeownership gap of 50\% among young adults, following the  methodology of \cite{Ashman2020} and \cite{aliprantis2022dynamics}. I find that parental transfers account for 11 percentage points (20\%) of the Black-White homeownership gap.

@@

This lower cost of illiquidity for households with wealthy parents offers a natural explanation for the positive empirical relationship between parental wealth and successful homeownership I found in the PSID. It also aligns with \citet{choukhmane2023benefits}, who find that households with wealthy parents are more likely to invest in illiquid retirement accounts and avoid costly early withdrawals, even after controlling for a wide set of characteristics.

@@

My paper first contributes to the literature studying the determinants of homeownership over the life cycle. Recent papers focus on marriage and family formation \citep{Fisher2011,Chang2024,Khorunzhina2019}, housing demand in old age \citep{McGee2019,Barczyk2020a}, and changing borrowing constraints \citep{Paz-Pardo2019,Mabille2020}. These studies highlight the importance of credit constraints and minimum down payments in constraining demand for owner-occupied housing. By considering parent-child interactions, I show how parental wealth affects the relative importance of constraints. For households with poorer parents, the sales and purchase costs decrease ownership by making it riskier to own. This risk is smaller for households with richer parents, making mortgage credit constraints relatively more important. My results highlight that policies increasing ownership through subsidies to first-time buyers---a common approach in many cities---will lead to increased housing inequality. My model-based results also complement some of the empirical research on the effect of parental resources and housing outcomes \citep[see e.g.,][ for recent work]{wold2024housing,daysal2023intergenerational,benetton2022dynastic,boileau2023drives} by quantifying the contribution of transfers to homeownership.

My paper also contributes to the literature on altruistic households interacting without commitment by studying how illiquid housing affects the commitment problem. In these models, children have an incentive to undersave to increase parental transfers (e.g., \citealp{Altonji1997a,Boar2018,Barczyk2014,Chu2020}). Illiquid housing imposes future expenditure commitments \citep{Chetty2007,Shore2010}, which children of wealthy parents strategically exploit to encourage future parental transfers. The paper most closely related to mine is \cite{Barczyk2020a}, who study how the homeownership of retired parents affects the economic behavior of the elderly and their adult children. In their model, the parents' home serves as a commitment device, encouraging wealth bequests from the parent and informal care from the adult offspring. My contribution is to study the complementary problem: how parental transfers affect the homeownership decisions of young adults. Finally, while \cite{kaplan2012moving} studies cohabitation with parents among individuals under the age of 25, my focus is on homeownership decisions for households from age 25 onwards.

@@

\subsection{Parental Transfers for Downpayments Over Time}\label{sec:overtime}
I use the Survey of Household Economics and Decisionmaking (SHED) and American Housing Survey (AHS) datasets to show the time trends in parental housing transfers. A benefit of these surveys is that they ask questions directly relating transfers to downpayments, unlike the PSID.

@@

	 {\begin{footnotesize} \textit{Notes:} The left panel uses data from the SHED with the sample restricted to current owners who report this being their first home. The right panel uses data from the AHS with the sample restricted to all owner-occupied units that did not indicate the sale of a previous home as the main source of funding.  Note that the left figure has year of purchase on the horizontal axis while the right figure has survey year. Dashed lines denote 95\% confidence intervals.\end{footnotesize}}
\end{figure}

@@

@@

\textit{Matching Parents with Children:} I use the Parent/Child file from the PSID's 2013 transfer supplement. I can observe each household's reported transfers to and from parents and children. This leads to discrepancies, where the child and parent do not agree on the amount given from the parent to the child. For each child, I use the parent's reported transfer to that child. First, there may be some stigma about receiving transfers, which may induce receiving children to underreport. Second, in the model, parents determine the size of the transfer they give to their child.

\textit{Definition of Transfers:} To measure the transfer receipt rate, I rely on the transfer supplement which asked all parents whether they gave money, gifts, or loans of \$100 or more to each child in 2012. I follow the literature \citep[e.g.,][]{mcgarry2016dynamic} and treat all transfers as gifts. I ignore transfers below \$500 to focus on transfers that are quantitatively meaningful for housing. About 25\% of transfers are below this threshold, and ignoring them increases the conditional mean transfer from \$2,921 to \$3,960 and reduces the transfer receipt rate from 45\% to 22\%. It is unsurprising that PSID-based transfer rates exceed those in the SCF \citep[e.g.,][]{feiveson2019lifecycle,sabelhaus2024household}, which focuses on substantial transfers. In the United States, inter-vivos transfers of the magnitudes observed here are far below the federal gift tax limits and are therefore untaxed for both giver and recipient. In 2024, gifts of up to \$18{,}000 per recipient require no filing; gifts above that require a gift-tax return, but tax is owed only if cumulative lifetime gifts and bequests exceed \$13.61 million.

In all other waves, there is a general question of gifts and inhertitances from people outside the household. We do not know the source of these transfers (e.g., they could be from friends or grandparents) and hence these do not know the extent to which these captures parental assistance. Nevertheless, they still capture some of the impact of transfer receipt on household decisions. I therefore include these variables in some of the empirical results, following \citet{Lee2018}.

Finally, why focus on inter-vivos transfers when most intra-family wealth transfers occur through bequests \citep[see e.g.,][]{feiveson2019lifecycle}? First, parental bequests are typically received after age 45, too late to affect young adults' transition into homeownership. Second, some bequests are given early and recorded as inter-vivos transfers. Third, inter-vivos transfers are often targeted to children's immediate liquidity needs, such as financing a home purchase, whereas bequests are often passive outcomes of unspent wealth. Fourth, while bequests dominate in aggregate terms, they are highly concentrated among wealthy households, whereas inter-vivos transfers are the main form of transfer for most parents \citep{yang2023financial}. Thus, inter-vivos transfers are more relevant for understanding early-life economic outcomes.

@@

The first two columns of the top panel compare households by transfers receipt. The mean transfer is relatively high at \$3,960, and 21\% of households received transfers in the last year. The largest difference is parental wealth: transferring parents are 2.5 times as wealthy as nontransferring parents. Perhaps surprisingly, receivers are slightly richer than nonreceivers (\$108,000 vs. \$78,500), are more likely to be college-educated and white, and are one year younger. Receivers are slightly less likely to be homeowners (39\% vs. 43\%), reflecting the age difference and the tendency for college attendance to delay homeownership.

@@

\subsubsection{Transfer Sizes}
For additional detail on the transfer variables, I plot kernel density estimates in Figure \ref{fig:PSID_transfer_to_renters}, for renters in $t$, separated by whether they own or rent in $t+2$, for the two transfer definitions. There are two takeaways. First, households who transition to ownership receive larger transfers on average, under both transfer measures. Second, the transfers in the transfer supplement are much smaller than than those in the gift/inheritance question. Indeed, the median parental transfer is about \$1k, the mean about \$3k, and the 95th percentile is at \$10k, compared to the median gift/inheritance of \$25k, mean of \$60k, and a 95th percentile of about \$130k.

@@

	 {\begin{footnotesize} \textit{Notes:} The left panel uses the transfers, gifts, and inheritance variable which is available in all waves of the PSID while the right panel uses the parental transfer data only available in the 2013 transfer supplement. See main text for a discussion. Vertical dotted and dashed lines denote the median and means, respectively. All estimates are weighted and omit the top 5\% of transfer sizes. \end{footnotesize}}
\end{figure}

While the PSID allows us to observe transfers and changes in housing tenure over time, it has several limitations. First, the transfer supplement captures transfers only in 2012, while the PSID is collected biennially, leading to undercounts of both transfer rates and amounts. Second, in the gift/inheritance question, the source of the transfer is unknown. Third, it is unclear whether intra-family loans are consistently reported or repaid.

@@

I now utilize microdata from the PSID to examine the empirical association between parental financial resources on housing outcomes. The sample is limited to households aged 25 to 44 to focus on the housing choices of young adults. I use the sum of parental net worth and income (denoted as wealth) to measure parents' financial resources in all specifications for two reasons. First, this is consistent with models of altruism, as the one used later in the paper. Second, since parents' wealth and income are highly correlated and I have few observations---fewer than 1,000 for many specifications---the standard errors are large when both are included.

\subsubsection{Transfers and the Transition into Homeownership}
Since many papers have already estimated the relationship between parental transfers and transitions into homeownership (e.g.\ \citealp{wold2024housing}; \citealp{Blickle2019}; \citealp{benetton2022dynastic}; \citealp{Lee2018}), I replicate these regressions to highlight the association between transfer receipt and rent-to-own transitions, but relegate the full results to Appendix~\ref{app:rent_to_own}. Receiving a parental transfer above \$500 is associated with a 1.2 percentage point increase in the probability of becoming a homeowner (not statistically significant), after controlling for household characteristics. Receiving large transfers above \$10,000 is associated with a 7.9 percentage point increase in the probability of becoming a homeowner ($p<0.1$), even after including household fixed effects. These are sizable effects relative to the baseline rent-to-own transition probability of about 18\% during the sample period.

@@

First, the left panel of Figure \ref{fig:owner_hazard} plots the share of owners aged 25 to 44 in year $t$ who remain owners in period $t+\Delta$. Among first-time owners, only 80\% have retained ownership four years later. For non-first-time owners, there is still a marked decrease though it is more muted, after four years about 10\% of all owners will be renting.

To understand which variables predicts retainment of ownership, I estimate
\begin{equation}	\label{eq:regmaintain}
	\Pr\left(Own_{i,t+\Delta}=1|Own_{i,t}=1\right)
	= \beta\ln(NetWorth_{i,t}) + \zeta\ln(ParWealth_{i,t})
	  + \gamma' X_{i,t} + \varepsilon_{i,t},
\end{equation}
where $Own_{i,t+\Delta}=1$ denote ownership at time $t+\Delta$, $\beta$ and $\zeta$ measure the effects of log household net worth and parental wealth, respectively. The vector of control variables  $X_{i,t}$ consists of log household income, log household size, age, parental age, and dummies for  education, marital status, race, household composition changes, and state and year fixed effects.  The sample is limited to households aged 25 to 44. These regressions are motivated by \cite{bond2021role}, who estimate similar equations in the PSID and the Health and Retirement Study. Figures \ref{fig:coef_allowner} and \ref{fig:coef_ftowner} plot the two coefficients of interest over different time horizons.

@@

I first limit the sample to first-time buyers to focus on distress among new owners. In a univariate regression, a 10\% increase in parental wealth is associated with a 0.07 percentage point decrease in the probability of being behind on mortgages two years later ($p<0.1$) With the full set of controls, the effect increases slightly, though it is no longer statistically significant. These magnitudes are economically meaningful since 1.6\% of first-time owners are behind on their mortgage. In Columns 3 to 5, I expand the analysis to include all owners who are currently not behind, increasing the sample sizes and allowing us to include household fixed effects. In these specifications, a 10\% increase in parental wealth is associated with a 0.09 percentage point decrease in delinquency risk in the univariate OLS regression ($p<0.001$), a 0.02 percentage point decrease in the multivariate OLS regression (not significant), and a 0.07 percentage point decrease in the fixed effects regression ($p<0.1$).

@@

	\end{threeparttable}
	{\begin{footnotesize}\begin{flushleft}
		\textit{Notes:} The estimated association between parental wealth and the probability of behind on mortgage payments in the next period from regressions using specification \ref{eq:regbehind}. See Appendix Table~\ref{tab:hypo_long} for all results. Standard errors in parentheses, clustered at the household level. \textsuperscript{+}$p<0.10$, \textsuperscript{*}$p<0.05$, \textsuperscript{**}$p<0.01$, \textsuperscript{***}$p<0.001$.
		\end{flushleft}\end{footnotesize}}	
\end{table}

@@

\subsubsection{Owners with Wealthy Parents Don't Downsize in Unemployment}\label{sec:eventstudy}
Distressed and forced home sales are costly and often triggered by unemployment \citep{kermani2021racial,hsu2018unemployment}. To test whether parental wealth mitigates this risk, I perform a simple event study of how unemployment affects the likelihood of downsizing.

The exercise follows \cite{Chetty2007} closely. The outcome of interest is the change in log housing consumption, which is set to zero for households who do not move. For movers, housing consumption is defined as yearly rent while renting and as imputed rent for owners, set to 5\% of the market value \citep{Davis2008}. The sample is limited to household heads who experience only one unemployement spell between ages 25 and 44. I divide the sample by whether a household's parents were in the top parental wealth quartile at the time of unemployment and compare the two groups' housing consumption growth at unemployment. The results are displayed in Figure \ref{fig:housinggrowthrates}. Among households with non-wealthy parents, housing consumption falls by about 5\% during unemployment ($p<0.05$). In contrast, households with wealthy parents see a slight increase (not significant) in housing consumption during unemployment. Appendix~\ref{app:eventstudy_details} shows that these patterns are robust to an event-study specification that controls for household characteristics. I return to these results later as part of the model validation.

@@

	{\begin{footnotesize} \textit{Notes:} Solid lines denote means, and bars denote the 95\% confidence interval. The sample consists of households aged 25-45 with exactly one unemployment spell and without changes in head and/or spouse in the four years before and after unemployment. \end{footnotesize}}
\end{figure}

>>>>

\usepackage[skip=4pt]{caption}    % space between caption and body (default ~10pt)
\setlength{\textfloatsep}{10pt}    % text↔float (above & below)  [default ~20pt]
% \setlength{\intextsep}{12pt}       % inline floats (here/ht)     [default ~12–20pt]
\newcommand{\E}{\mathbb{E}}
\newcommand{\mc}{\mathcal}

\title{{\LARGE Illiquid Homeownership and the \\ Bank of Mom and Dad}\thanks{Federal Reserve Board of Governors. E-mail: eirik.e.brandsaas@frb.gov. I have benefited from the comments of Dean Corbae, Erwan Quintin, Ananth Seshadri, Corina Mommaerts, Corina Boar, Manuel Amador, Hamish Low, Patrick Moran, Dionissi Aliprantis, David Sturrock, Mike Eriksen, Aradhya Sood, Margaret Jacobson, William Gamber, Claudia Sahm, John Sabelhaus, and Joel McMurry, as well as seminar participants Oxford Sa{\"i}d, Federal Reserve Board, Cal State Fullerton, Boston Fed, ECARES/Université libre de Bruxelles, Santa Clara, University of Lugano, NYU Furman, Norges Bank, BI Norwegian Business School, UW-Madison, and conference participants at NASEMS, AREUEA National, ReCapNet, Minneapolis Fed/OIGI, UEA EU, CFPB, CEF, SAET, Oxford NuCamp Macro Workshop, YES 2020 at UPenn, and Minnesota-Wisconsin Macro/International Workshop. Code and replication instructions are available at the \href{https://github.com/eirikbrandsaas/HomeownershipBankMomDad.jl}{author's  website and Github}.\newline \textit{Disclaimer: }The views expressed in this paper are solely the responsibility of the author and should not be interpreted as reflecting the views of the Board of Governors of the Federal Reserve System or of anyone else associated with the Federal Reserve System.}}
\author{Eirik Eylands Brandsaas}
\date{{\today }}

@@

\maketitle
\begin{abstract}
{\begin{spacing}{1.2}
Housing is the largest asset in U.S. household portfolios, and first-time homebuyers increasingly rely on parental transfers. This paper quantifies the contribution of parental transfers to the homeownership rate of young households. I build and estimate a life-cycle overlapping generations model with housing, where adult children and parents interact without commitment. I find that parental transfers account for 13 percentage points (27\%) of young households' homeownership. Transfers from wealthy parents not only help households overcome borrowing constraints, but also help sustain homeownership, mitigating the drawbacks of illiquidity. Surprisingly, policies lowering entry barriers to homeownership generally increase the reliance on parental wealth, whereas increased liquidity reduces it. Finally, I show that children of wealthy parents strategically use the illiquidity of housing as a commitment device to encourage transfers, resulting in a preference for illiquidity.
\end{spacing}}
\vspace{0.5cm}
\noindent \textbf{JEL:} D14, D15, E21, G51, R21

@@

\section{Introduction}
Housing is the largest asset in U.S. household portfolios. While homeownership is seen as central to the ``American Dream'' \citep{goodman2018homeownership}, households face different barriers to ownership. Between 2009 and 2016, about 30\% of first-time buyers received direct parental help with the down payment (Survey of Household Economics and Decisionmaking; see Section~\ref{sec:data}), highlighting disparities in access by parental wealth.

@@

To quantify the impact of parental transfers on homeownership, I then combine two distinct models: housing models, which ignore parental transfers, and models of parental transfers, which ignore housing. I build a life-cycle overlapping generations model with altruistic parents and a rent-or-own decision. Parents and children interact without commitment through both inter-vivos transfers and end-of-life bequests.  In altruism models without housing (e.g., \citealp{Altonji1997a,Barczyk2014}), transfers generally flow to borrowing-constrained children with high marginal utility of wealth. Introducing a frictional housing market adds two additional channels. First, transfers can directly alleviate credit constraints, enabling wealthier parents to help their children meet down-payment requirements. Second, future transfers provide partial insurance against future income shocks, reducing risks associated with the large illiquid investment of homeownership. Moreover, the illiquidity of housing serves as a commitment device enabling children to secure future transfers. These mechanisms differ from those examined by \citet{Barczyk2020a}, who study how elderly parents’ homeownership affects nursing-home entry, care arrangements, and transfers.  As in all transfer models, anticipated transfers crowd out children’s saving, potentially lowering homeownership in isolation.

@@

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Study of Income Dynamics (PSID). To quantify the importance of transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 13 percentage points (27\%) of the homeownership rate among young households. In one application of my model, I quantify the contribution of parental transfers to the Black-White homeownership gap of 50\% among young adults, following the  methodology of \cite{Ashman2020} and \cite{aliprantis2022dynamics}. I find that parental transfers account for 11 percentage points (20\%) of the Black-White homeownership gap. On the housing side, the baseline is intentionally parsimonious---a simple rent-or-own choice with only two sizes, a fixed rent–price ratio, and prices held constant---to isolate the new parent–child channel.

@@

This lower cost of illiquidity for households with wealthy parents offers a natural explanation for the positive empirical relationship between parental wealth and successful homeownership I find in the PSID. It also aligns with \citet{choukhmane2023benefits}, who find that households with wealthy parents are more likely to invest in illiquid retirement accounts and avoid costly early withdrawals, even after controlling for a wide set of characteristics.

@@

My paper first contributes to the literature studying the determinants of homeownership over the life cycle. Recent papers focus on marriage and family formation \citep{Fisher2011,Chang2024,Khorunzhina2019}, housing demand in old age \citep{McGee2019,Barczyk2020a}, and changing borrowing constraints \citep{Paz-Pardo2019,Mabille2020}. These studies highlight the importance of credit constraints and minimum down payments in constraining demand for owner-occupied housing. By considering parent-child interactions, I show how parental wealth affects the relative importance of constraints. For households with poorer parents, the sales and purchase costs decrease ownership by making it riskier to own. This risk is smaller for households with richer parents, making mortgage credit constraints relatively more important. My results highlight that policies increasing ownership through subsidies to first-time buyers---a common approach in many cities---will lead to increased housing inequality. My model-based results also complement some of the empirical research on the effect of parental resources and housing outcomes \citep[see e.g.,][for recent work]{wold2024housing,daysal2023intergenerational,benetton2022dynastic,boileau2023drives} by quantifying the contribution of transfers to homeownership.

My paper also contributes to the literature on altruistic households interacting without commitment by studying how illiquid housing affects the commitment problem. In these models, children have an incentive to undersave to increase parental transfers (e.g., \citealp{Altonji1997a,Boar2018,Barczyk2014,Chu2020}). Illiquid housing imposes future expenditure commitments \citep{Chetty2007,Shore2010}, which children of wealthy parents strategically use to encourage future parental transfers. The paper most closely related to mine is \cite{Barczyk2020a}, who study how the homeownership of retired parents affects the economic behavior of the elderly and their adult children. In their model, the parents' home serves as a commitment device, encouraging wealth bequests from the parent and informal care from the adult offspring. My contribution is to study the complementary problem: how parental transfers affect the homeownership decisions of young adults. Finally, while \cite{kaplan2012moving} studies cohabitation with parents among individuals under the age of 25, my focus is on homeownership decisions for households from age 25 onward.

@@

\subsection{Parental Transfers for Down Payments Over Time}\label{sec:overtime}
I use the Survey of Household Economics and Decisionmaking (SHED) and American Housing Survey (AHS) to show the time trends in parental housing transfers. Unlike the PSID, these surveys ask questions directly relating transfers to down payments.

@@

	\caption*{\begin{footnotesize} \textit{Notes:}  Left panel (SHED): sample is current owners reporting this is their first home; x-axis is purchase year; outcome is the share receiving any down-payment transfer. Right panel (AHS): sample is owner-occupied units that do not report proceeds from a prior home sale as the main funding source; x-axis is survey year; outcome is the share whose down payment was mostly funded by a transfer. Dashed lines denote 95\% confidence intervals.\end{footnotesize}}
\end{figure}

@@

Unfortunately, the Health and Retirement Study (HRS)---though it collects transfer data in many waves---does not include detailed information on child household wealth, a key determinant of homeownership. I therefore rely on the PSID, which includes detailed household-level data for both generations.

@@

\textit{Matching Parents with Children:} I use the Parent/Child file from the PSID's 2013 transfer supplement. I can observe each household's reported transfers to and from parents and children. This leads to discrepancies, where the parent and child do not agree on the amount given from the parent to the child. For each child, I use the parent's reported transfer to that child. First, there may be some stigma about receiving transfers, which may induce receiving children to underreport. Second, in the model, parents determine the size of the transfer they give to their child.

\textit{Definition of Transfers:} To measure the transfer receipt rate, I rely on the transfer supplement, which asked all parents whether they gave money, gifts, or loans of \$100 or more to each child in 2012. I follow the literature \citep[e.g.,][]{mcgarry2016dynamic} and treat all transfers as gifts. I ignore transfers below \$500 to focus on transfers that are quantitatively meaningful for housing. About 25\% of transfers are below this threshold, and ignoring them increases the conditional mean transfer from \$2,921 to \$3,960 and reduces the transfer receipt rate from 45\% to 22\%. It is unsurprising that PSID-based transfer rates exceed those in the SCF \citep[e.g.,][]{feiveson2019lifecycle,sabelhaus2024household}, which focuses on substantial transfers. In the United States, inter-vivos transfers of the magnitudes observed here are far below the federal gift tax limits and are therefore untaxed for both giver and recipient. In 2024, gifts of up to \$18{,}000 per recipient require no filing; gifts above that require a gift-tax return, but tax is owed only if cumulative lifetime gifts and bequests exceed \$13.61 million.

In all other waves, there is a general question on gifts and inheritances from people outside the household. We do not know the source of these transfers (e.g., they could be from friends or grandparents), and hence we do not know the extent to which they capture parental assistance. Nevertheless, they still capture some of the impact of transfer receipt on household decisions. I therefore include these variables in some of the empirical results, following \citet{Lee2018}.

Finally, why focus on inter-vivos transfers when most intrafamily wealth transfers occur through bequests \citep[see e.g.,][]{feiveson2019lifecycle}? First, parental bequests are typically received after age 45, too late to affect young adults' transition into homeownership. Second, some bequests are given early and recorded as inter-vivos transfers. Third, inter-vivos transfers are often targeted to children's immediate liquidity needs, such as financing a home purchase, whereas bequests are often passive outcomes of unspent wealth. Fourth, while bequests dominate in aggregate terms, they are highly concentrated among wealthy households, whereas inter-vivos transfers are the main form of transfer for most parents \citep{yang2023financial}. Thus, inter-vivos transfers are more relevant for understanding early-life economic outcomes.

@@

The first two columns of the top panel compare households by transfer receipt. The mean transfer is relatively high at \$3,960, and 21\% of households received transfers in the last year. The largest difference is parental wealth: transferring parents are 2.5 times as wealthy as nontransferring parents. Perhaps surprisingly, receivers are slightly richer than nonreceivers (\$108,000 vs. \$78,500), are more likely to be college-educated and white, and are one year younger. Receivers are slightly less likely to be homeowners (39\% vs. 43\%), reflecting the age difference and the tendency for college attendance to delay homeownership.

@@

\subsubsection{Transfer Sizes}
For additional detail on the transfer variables, I plot kernel density estimates in Figure \ref{fig:PSID_transfer_to_renters}, for renters in $t$, separated by whether they own or rent in $t+2$, for the two transfer definitions. There are two takeaways. First, households who transition to ownership receive larger transfers on average, under both transfer measures. Second, the transfers in the transfer supplement are much smaller than than those in the gift/inheritance question. Indeed, the median parental transfer is about \$1{,}000, the mean about \$3{,}000, and the 95th percentile is about \$10{,}000, compared to the median gift/inheritance of \$25{,}000, mean of \$60{,}000, and a 95th percentile of about \$130{,}000.

@@

	 \caption*{\begin{footnotesize} \textit{Notes:} The left panel uses the transfers, gifts, and inheritances variable, which is available in all waves of the PSID, while the right panel uses the parental-transfer data only available in the 2013 transfer supplement. See main text for a discussion. Vertical dotted and dashed lines denote the median and means, respectively. All estimates are weighted and omit the top 5\% of transfer sizes. \end{footnotesize}}
\end{figure}

While the PSID allows us to observe transfers and changes in housing tenure over time, it has several limitations. First, the transfer supplement captures transfers only in 2012, while the PSID is collected biennially, leading to undercounts of both transfer rates and amounts. Second, in the gift/inheritance question, the source of the transfer is unknown. Third, it is unclear whether intrafamily loans are consistently reported or repaid.

@@

I now utilize microdata from the PSID to examine the empirical association between parental financial resources and housing outcomes. The sample is limited to households aged 25 to 44 to focus on the housing choices of young adults. I use the sum of parental net worth and income (denoted as wealth) to measure parents' financial resources in all specifications for two reasons. First, this is consistent with models of altruism, such as the one used later in the paper. Second, since parents' wealth and income are highly correlated and I have few observations---fewer than 1,000 for many specifications---the standard errors are large when both are included. Note that neither transfer receipt nor parental wealth is treated as exogenous. For example, many transfers may be made precisely because the renter is buying a home, or parents may save in order to help their children enter the housing market.

\subsubsection{Transfers and the Transition into Homeownership}
Since many papers have already estimated the relationship between parental transfers and transitions into homeownership (e.g., \citealp{wold2024housing}; \citealp{Blickle2019}; \citealp{benetton2022dynastic}; \citealp{Lee2018}), I replicate these regressions to highlight the association between transfer receipt and rent-to-own transitions, but relegate the full results to Appendix~\ref{app:rent_to_own}. Receiving a parental transfer above \$500 is associated with a 1.2 percentage point increase in the probability of becoming a homeowner (not statistically significant), after controlling for household characteristics. Receiving large transfers above \$10,000 is associated with a 7.9 percentage point increase in the probability of becoming a homeowner ($p<0.1$), even after including household fixed effects. These are sizable effects relative to the baseline rent-to-own transition probability of about 18\% during the sample period.

@@

First, the left panel of Figure \ref{fig:owner_hazard} plots the share of owners aged 25 to 44 in year $t$ who remain owners in period $t+\Delta$. Among first-time owners, only 80\% have retained ownership four years later. For non-first-time owners, there is still a marked decrease, though it is more muted; after four years about 10\% of all owners will be renting.

To understand which variables predicts retention of ownership, I estimate
\begin{equation}	\label{eq:regmaintain}
	\Pr\left(Own_{i,t+\Delta}=1|Own_{i,t}=1\right)
	= \beta\ln(NetWorth_{i,t}) + \zeta\ln(ParWealth_{i,t})
	  + \gamma' X_{i,t} + \varepsilon_{i,t},
\end{equation}
where $Own_{i,t+\Delta}=1$ denotes ownership at time $t+\Delta$, $\beta$, and $\zeta$ measure the effects of log household net worth and parental wealth, respectively. The vector of control variables  $X_{i,t}$ consists of log household income, log household size, age, parental age, and dummies for  education, marital status, race, household composition changes, and state and year fixed effects.  The sample is limited to households aged 25 to 44. These regressions are motivated by \cite{bond2021role}, who estimate similar equations using the PSID and the HRS. Figures \ref{fig:coef_allowner} and \ref{fig:coef_ftowner} plot the two coefficients of interest over different time horizons.

@@

I first limit the sample to first-time buyers to focus on distress among new owners. In a univariate regression, a 10\% increase in parental wealth is associated with a 0.07 percentage point decrease in the probability of being behind on mortgages two years later ($p<0.1$). With the full set of controls, the effect increases slightly, though it is no longer statistically significant. These magnitudes are economically meaningful since 1.6\% of first-time owners are behind on their mortgage. In Columns 3 to 5, I expand the analysis to include all owners who are currently not behind, increasing the sample sizes and allowing us to include household fixed effects. In these specifications, a 10\% increase in parental wealth is associated with a 0.09 percentage point decrease in delinquency risk in the univariate OLS regression ($p<0.001$), a 0.02 percentage point decrease in the multivariate OLS regression (not significant), and a 0.07 percentage point decrease in the fixed-effects regression ($p<0.1$).

@@

	\end{threeparttable}
	{\begin{footnotesize}\begin{flushleft}
		\textit{Notes:} Estimated association between parental wealth and the probability of being behind on mortgage payments in the next period from regressions using specification \ref{eq:regbehind}. See Appendix Table~\ref{tab:hypo_long} for all results. Standard errors in parentheses, clustered at the household level. \textsuperscript{+}$p<0.10$, \textsuperscript{*}$p<0.05$, \textsuperscript{**}$p<0.01$, \textsuperscript{***}$p<0.001$.
		\end{flushleft}\end{footnotesize}}	
\end{table}

@@

\subsubsection{Owners with Wealthy Parents Don't Downsize in Unemployment}\label{sec:eventstudy}
Because unemployment often triggers costly distressed sales \citep{kermani2021racial,hsu2018unemployment}, I test whether parental wealth mitigates this risk by estimating an event study of unemployment’s effect on downsizing.

The exercise follows \citet{Chetty2007}. The outcome is the change in log housing consumption (winsorized at the 1st and 99th percentiles), set to zero for non-movers. For movers, housing consumption equals annual rent if renting and imputed rent if owning (5\% of market value; \citealp{Davis2008}). The sample includes household heads with exactly one unemployment spell between ages 25 and 44. I split the sample by whether parental wealth in the unemployment year lies in the top quartile and compare housing-consumption growth around unemployment. The results are displayed in Figure~\ref{fig:housinggrowthrates}. Among households with non-wealthy parents, housing consumption falls by about 5\% during unemployment ($p<0.05$). In contrast, households with wealthy parents exhibit a small, statistically insignificant increase. Appendix~\ref{app:eventstudy_details} gives additional details and shows that these patterns are robust to an event-study specification with household controls and to alternative cutoffs. I return to these results in the model validation.

@@

	\caption*{\begin{footnotesize} \textit{Notes:} Solid lines denote means, and bars denote 95\% confidence intervals. The sample consists of households aged 25-45 with exactly one unemployment spell and without changes in head and/or spouse in the four years before and after unemployment. \end{footnotesize}}
\end{figure}

\subsection{Taking Stock of the Data}
Taking stock, parental transfers have become more important for young homeowners over the past two decades. About 20\% of young households receive transfers from their parents in a given year, and recipients have wealthier parents. Recipients are more likely to transition from renting to owning, after conditioning on a rich set of covariates. Parental wealth is also positively associated with maintaining homeownership. Households with richer parents are less likely to fall behind on mortgages and to downsize during unemployment spells. These results suggest that homeowners with wealthy parents find the illiquidity of homeownership less problematic.
